Cultural Transmission and Technological Transitions during the Late Paleolithic in Korea

### 2023-02-09 analysis/paper/paper.qmd
brainstorming for discussion

Following @bettinger1999point we computed Pearson's correlation coefficients for all attributes on the stemmed points. The correlation of artifact attributes was proposed by @bettinger1999point to be a key indicator of indirect bias during cultural transmission. This is based on the assumption that artifact-making was transmitted in packages of traits taken from socially successful individuals who were role-models. In their work, @bettinger1999point interpreted correlation coefficient values of around 0.5 and higher as evidence of indirect bias, and lower values as evidence of guided variation. In this research, we examine correlation between attributes and explore other variables such as different phases, raw materials and assemblages.

@@

###### facet plot
cv_plot_raw_materials_label %>%
  ggplot(aes(x = variable, y = cv_by_materials)) +
  geom_point() +
  geom_errorbar(aes(ymin = cv_low_sharma,
                    ymax = cv_high_sharma), width = .2) +
    # draw a line to show our threshold value
  geom_hline(yintercept = 25, colour = "red") +
  xlab("Raw Materials") +
  ylab("Coefficient of Variation on Attributes") +
  facet_wrap( ~ label, scales = "free_y") +
  coord_cartesian(ylim = c(0,250)) +
  theme_bw() 
```

# Discussion

To investigate the social contexts of technological transitions represented by the emergence of stemmed points during the Korean Late Paleolithic, we applied concepts of cultural transmission. We asked three questions to examine the dominant mode of cultural transmission and the transmission process over time: what is the dominant mode of cultural transmission for technological innovation in the Korean Late Paleolithic? Do the modes of cultural transmission vary over time? and Do the modes of cultural transmission vary over space? We focused on two transmission biases: guided variation (socially isolated groups made stemmed points through trial and error) and indirect bias (socially connected groups whose knowledge of stemmed points derived from copying others). We set roughly `r threshold_value` as our threshold of the CV value to decide between these two types of transmission biases, based on prior work on Great Basin Projectiles [@eerkens2001techniques]. If the CV value is lower than 25, we interpreted this as indicating indirect bias while values higher than 25 were interpreted as guided variation.

The CV values for all stemmed point attributes are mostly over 25, except for maximum length, which is just below. Change in CV values for over time in the Korean Late Paleolithic period is complex, with no clear directional changes (@fig-plot-cv-over-time). Comparing chronological phases 2 and 3, we found that there are only minor differences between the two phases. Applying the Modified Signed-Likelihood Ratio Test (SLRT) <!-- GP to add a citation for this test here --> to test for the equality of CVs, we found no statistically significant changes in the CV values of any attributes (@tbl-cv-over-time) between the two phases. In exploring individual assemblages that contain multiple stemmed points we found confirmation of minimal change in CV values over time (@fig-plot-cv-four-assemblage).

@@

In exploring geographical variation we found no clear pattern variation in CV between sites with more than five stemmed points (@fig-plot-cv-by-site). The most striking pattern in this site comparison is that CV values of body length, maximum length, mid width, and tang width are generally lower than tang length, stem length, and stem width, similar to what is evident in @fig-plot-cv-all-attributes. The attributes with lower CV values could be transmitted in a social context dominated by indirect bias, while the higher CV attributes could be influenced more by specific manufacturing and maintenance situations. We assume that the lower CV values of the four attributes could imply that those four attributes are closely associated with the projectile function of the tool so that the knappers wanted to standardize the shape. For example, body length is related to penetration, durability, hardness, and rejuvenation of a projectile point and tang width is connected to wound damage, penetration, durability and hardness [@odell1986experiments; @shea2001experimental; @cheshier2006projectile; @yaroshevich2016projectile; @bebber2017developing; @wood2018wound]. This can be contrasted with tang length, stem length, and stem width, which are away from the impact end of the artifact. These were allowed to vary more freely between points, perhaps to be able to accommodate shafts of different types of wood with varying properties of strength, flexibility, and weight.

@@

<!-- BM: GP to write up the discussion of the correlation analysis in detail. Use this paper as a template for how detailed you need to be here: "Point Typologies, Cultural Transmission, and the Spread of Bow-and-Arrow Technology". Aim to improve your writing, make it better than your usual style. -->

>>>>

Following @bettinger1999point we computed Pearson's correlation coefficients for all attributes on the stemmed points. The correlation of artifact attributes was proposed by @bettinger1999point to be a key indicator of indirect bias during cultural transmission. This is based on the assumption that artifact-making was transmitted in packages of traits taken from socially successful individuals who were role-models. In their work, @bettinger1999point interpreted correlation coefficient values of around `r threshold_value_cor` and higher as evidence of indirect bias, and lower values as evidence of guided variation. In this research, we examine correlation between attributes and explore other variables such as different phases, raw materials and assemblages.

@@

###### facet plot
cv_plot_raw_materials_label %>%
  ggplot(aes(x = variable, y = cv_by_materials)) +
  geom_point() +
  geom_errorbar(aes(ymin = cv_low_sharma,
                    ymax = cv_high_sharma), width = .2) +
    # draw a line to show our threshold value
  geom_hline(yintercept = 25, colour = "red") +
  xlab("Raw Materials") +
  ylab("Coefficient of Variation on Attributes") +
  facet_wrap( ~ label, scales = "free_y") +
  coord_cartesian(ylim = c(0,250)) +
  theme_bw() 
```
# Discussion

To investigate the social contexts of technological transitions represented by the emergence of stemmed points during the Korean Late Paleolithic, we applied concepts of cultural transmission. We asked three questions to examine the dominant mode of cultural transmission and the transmission process over time: what is the dominant mode of cultural transmission for technological innovation in the Korean Late Paleolithic? Do the modes of cultural transmission vary over time? and Do the modes of cultural transmission vary over space? We focused on two transmission biases: guided variation (socially isolated groups made stemmed points through trial and error) and indirect bias (socially connected groups whose knowledge of stemmed points derived from copying others). Following @bettinger1999point, we used `r threshold_value_cor` as a threshold of correlation coefficients to evaluate the two transmission biases. Correlation coefficients below `r threshold_value_cor` are indicative of guided variation and higher values indicate indirect bias. Since there is no clear criteria for CV, we set roughly `r threshold_value` as our threshold of the CV value to decide between these two types of transmission biases, based on prior work on Great Basin Projectiles [@eerkens2001techniques]. If the CV value is lower than `r threshold_value`, we interpreted this as indicating indirect bias while values higher than `r threshold_value` were interpreted as guided variation.

Our correlation analyses show that the correlation coefficients of stemmed points are all positive but mostly under `r threshold_value_cor`, showing the possibility of guided variation as their main transmission bias. Some correlations such as body length and maximum length have strong relationship while other correlations such as tang length and body length are much weaker. The correlation coefficients increase from chronological Phases 2 to 3 (Panel D of @fig-corr-analysis), but there is no clear pattern and the values vary depending on assemblages (@fig-plot-cor-four-assemblage). The CV values for all stemmed point attributes are mostly over `r threshold_value`, except for maximum length, which is just below. Change in CV values for over time in the Korean Late Paleolithic period is complex, with no clear directional changes (@fig-plot-cv-over-time). Comparing chronological phases 2 and 3, we found that there are only minor differences between the two phases. Applying the Modified Signed-Likelihood Ratio Test (MSLR) to test for the equality of CVs [@smallwood2022using; @krishnamoorthy2014improved], we found no statistically significant changes in the CV values of any attributes (@tbl-cv-over-time) between the two phases. In exploring individual assemblages that contain multiple stemmed points we found confirmation of minimal change in CV values over time (@fig-plot-cv-four-assemblage).

@@

```
In exploring geographical variation we found no clear pattern variation in both correlation coefficients (@fig-plot-cor-four-assemblage) and CV values (@fig-plot-cv-four-assemblage) between assemblages with more than five stemmed points. The four assemblages show different distribution patterns.

-SYG1_2 and SYG1_3 are from the same site but different time periods and the correlations decreases over time. SYG1_2 is from earlier period but having the highest median, 
-two groups of correlations in SYG1_2
-

The most striking pattern in this assemblage comparison is that CV values of body length, maximum length, mid width, and tang width are generally lower than tang length, stem length, and stem width, similar to what is evident in @fig-plot-cv-all-attributes. The attributes with lower CV values could be transmitted in a social context dominated by indirect bias, while the higher CV attributes could be influenced more by specific manufacturing and maintenance situations. We assume that the lower CV values of the four attributes could imply that those four attributes are closely associated with the projectile function of the tool so that the knappers wanted to standardize the shape. For example, body length is related to penetration, durability, hardness, and rejuvenation of a projectile point and tang width is connected to wound damage, penetration, durability and hardness [@odell1986experiments; @shea2001experimental; @cheshier2006projectile; @yaroshevich2016projectile; @bebber2017developing; @wood2018wound]. This can be contrasted with tang length, stem length, and stem width, which are away from the impact end of the artifact. These were allowed to vary more freely between points, perhaps to be able to accommodate shafts of different types of wood with varying properties of strength, flexibility, and weight.

@@
